---
title: "USA Inbound Border Crossing Data"
author: "Marc Alexander"
date: "2024-07-10"
output: 
  html_document: 

---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

  # knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
  # suppressWarnings({ })   

library(tidyr)
library(dplyr)
library(ggplot2)
library(broom)
library(reader)
library(scales)
library(hrbrthemes)
library(forcats)
library(mapproj)
library(gridExtra)
library(ggh4x)

getwd()
# setwd("~/Documents/Projects/usa-inbound_border_crossing")

# Loading the data from CSV file, 
data_00_base <- data.frame(reader("~/Documents/Projects/usa-inbound_border_crossing/data/Border_Crossing_Entry_Data.csv"))
# data_00_base <- data.frame(data_00_base)
# cleanse commas (e.g. 12,000) from the field that holds the measures numeric values
data_00_base$Value <- as.numeric(gsub("[^0-9.-]", "", data_00_base$Value))

# Add some features to the BASE data frame.  This is the BASE that all sub-queries (data frames) will pull from.
data_00_base_mod <- data_00_base %>%
  mutate(Border_Code = ifelse(grepl("Canada", Border, fixed = TRUE), "CAN", "MEX")) %>%
  separate(Date, c("Month", "Year"), remove = FALSE)  %>%
  # mutate(Measure = factor(Measure, order = TRUE, levels = c("Bus Passengers","Buses","Pedestrians","Personal Vehicle Passengers","Personal Vehicles","Rail Containers Empty","Rail Containers Loaded","Train Passengers","Trains","Truck Containers Empty","Truck Containers Loaded","Trucks"))) 
  # mutate(Measure = factor(Measure, order = TRUE, levels = c("Personal Vehicles","Personal Vehicle Passengers","Pedestrians","Trucks","Truck Containers Loaded","Truck Containers Empty","Buses","Bus Passengers","Trains","Train Passengers","Rail Containers Loaded","Rail Containers Empty"))) 
  mutate(measure_summary = factor(Measure, order = TRUE, levels = c("Personal Vehicles","Personal Vehicle Passengers","Pedestrians","Trucks","Truck Containers Loaded","Truck Containers Empty","Buses","Bus Passengers","Trains","Train Passengers","Rail Containers Loaded","Rail Containers Empty"))) %>%
  mutate(Measure = factor(Measure, order = TRUE, levels = c("Trains","Train Passengers","Rail Containers Loaded","Rail Containers Empty","Buses","Bus Passengers","Trucks","Truck Containers Loaded","Truck Containers Empty","Pedestrians","Personal Vehicles","Personal Vehicle Passengers"))) %>%
  # mutate(Year_num) = as.integer(Year)
  mutate(state_delete_me = State)
  
data_00_base_mod$Year_num <- as.integer(data_00_base_mod$Year)

  # Augment with a dimension table to help control colors and sorting in the presentation tier
  # This is all stuff I decided on when building the Tableau dashboard.  I want to follow that lead with R to make it easier to compare/contrast output side-by-side
  dim_measure <- data.frame(
    Measure=c("Bus Passengers","Buses","Pedestrians","Personal Vehicle Passengers","Personal Vehicles","Rail Containers Empty","Rail Containers Loaded","Train Passengers","Trains","Truck Containers Empty","Truck Containers Loaded","Trucks")
    ,mode_main_sort=c(8,7,3,2,1,11,10,12,9,6,5,4)
    ,mode_color_code=c("#C799BC","#8074A8","#4E79A7","#F59C3C","#C14F22","#CDCECD","#5B6570","#89C8CC","#848E93","#F4D166","#B2C25B","#34844A")
  )
  
  # Left Join the Base table with supplemental information about the Measures to help with the presentation layer
  data_00_base_mod <- merge(x = data_00_base_mod, y = dim_measure, by = "Measure", all.x = TRUE)
  
  # A little controller for the colors of the countries
  cols_border_crossing <- c("CAN" = "#F8756D", "MEX" = "#02BFC4")
  cols_world <- c("Canada" = "#F8756D", "Mexico" = "#02BFC4")
  cols_usa <- c("Canada" = "#F8756D", "Mexico" = "#02BFC4")
##########################################################################################################################################
# 10d for high-level Map graphs
##########################################################################################################################################
  
  ##########################################################################################################################################
  # Map Datasets
  world_map <- map_data("world") %>%
      filter(region == "Canada" | region == "Mexico" | region == "USA")
  
  us_map <- map_data("state")
  
  ##########################################################################################################################################
  # DATA: Border Crossing Dataset (aggregated to proper Grain)  
  data_10c_core <- data_00_base_mod %>%
    filter(measure_summary == "Personal Vehicles" | measure_summary == "Pedestrians" | measure_summary == "Trucks" | measure_summary == "Buses" | measure_summary =="Trains") %>% 
    filter(Year == "2023") %>%
    dplyr::select(Year, Border_Code, Latitude, Longitude, Port.Name, State, Value) %>%
    group_by(Year, Border_Code, Latitude, Longitude, Port.Name, State) %>%
    summarise(crossings = sum(Value), record_cnt = n(), .groups = 'keep') %>%
    mutate(size_class = cut(crossings,breaks=c(0,200000,400000,600000,1400000,Inf), labels=c("< 200K", "200K-400K", "400K-600K", "600K-1.4M", "1.4M-2.5M" ))) %>% 
    mutate(shape_class = cut(crossings,breaks=c(0,200000,400000,600000,1400000,Inf), labels=c(4,3,21,22,23))) %>%
    arrange(Border_Code, - crossings)  
  ##########################################################################################################################################
  # GRAPH: Border Crossing Dataset (aggregated to proper Grain)  
  suppressWarnings({ 
    gg_00a_map_intl_borders <- ggplot() + 
      geom_polygon(data=world_map, aes(x=long, y=lat, fill=region, group=group), color = "white", alpha = .9) + 
      geom_polygon(data=us_map, aes(x=long, y=lat, group=group), color = "white", alpha = 0, linewidth = .1) + 
      geom_point(data=data_10c_core, aes(x=Longitude, y=Latitude, size = size_class, stroke = .75), shape = 21, color = "black", fill="gray90") + 
      guides(fill="none") + 
      labs(title="North American Internationial Borders",
           subtitle='US International Borders and Ports of Entry',
           caption=c('Source: Bureau of Transportation Statistics'),
           size = c("2023 Inbound Crossings", "per Port of Entry")) +
        theme_minimal() +
        theme(
          axis.text.y = element_text(angle = 0),
          axis.title.y = element_blank(),
          panel.background = element_blank(),
          strip.background = element_rect(colour = "gray50", fill = "gray90"),
          panel.border = element_rect(colour = "gray50", fill = NA, size = .2), 
          panel.grid.minor.y = element_blank(),        
          panel.grid.major.x = element_blank(),        
          panel.grid.minor.x = element_blank()
        ) +
      # theme_void()+
      # theme(title=element_text(face='bold'),
      #       legend.position = 'right') + 
      # ggtitle('North America') + +
        labs(
          title = "USA International Borders | Mapped",
          # y = data_10d_core_melt$measure,
          subtitle = "Ports of Entry with the size of the mark mapped to # Border Crossings in 2023",
          caption = "Bureau of Transportation Statistics"
        ) +
      scale_colour_manual(values = cols_world) +
      coord_map(xlim = c(-160, -55), ylim = c(15, 68))    
  })
```

```{r echo=FALSE, fig.width = 10, fig.height=4}
##########################################################################################################################################
# Facet Wrap version of the Bar Graphs
##########################################################################################################################################    

##########################################################################################################################################
# DATA: Data gathering for 10d_core that will cover several high-level graphs

# Main Data Wrangling to get to the right section and grain
data_10d_core <- data_00_base_mod %>%
  filter(measure_summary == "Personal Vehicles" | measure_summary == "Pedestrians" | measure_summary == "Trucks" | measure_summary == "Buses" | measure_summary =="Trains") %>% 
  filter(Year == "2023") %>%
  dplyr::select(Year, Border_Code, State, Port.Name, Value) %>%
  group_by(Year, Border_Code) %>%
  summarise(ports_of_entry_cnt = n_distinct(Port.Name), crossings = sum(Value), states_on_border_cnt = n_distinct(State), .groups = 'keep') %>%
  mutate(border_miles = ifelse(grepl("CAN", Border_Code, fixed = TRUE), 5525, 1954)) 

# change the column order to be a little more readable (nice, but not required)
data_10d_core <- data_10d_core[,c(1,2,6,5,3,4)]

# Do some additional clean-up to the columns so we can cleanse after the column_names are melted to be rows
data_10d_core_rename <- data_10d_core
colnames(data_10d_core_rename) <- c("Year", "Border_Code", "Miles_of_Border", "States_on_Border", "Ports_of_Entry", "Border_Crossings")
data_10d_core_melt <- data_10d_core_rename %>%
  pivot_longer(!c(Year,Border_Code),names_to = "measure",values_to = "value") %>%
  mutate(measure = gsub("_"," ",measure)) %>%
  mutate(measure = factor(measure, order = TRUE, levels = c("Miles of Border", "Ports of Entry", "States on Border", "Border Crossings"))) %>%
  mutate(col_fill_color = as.factor(ifelse(grepl("CAN", Border_Code, fixed = TRUE), "#F8756D", "#33C051")))

  # Use this to create a custom discrete color scale that matches the map above
  cols_border_crossing<- c("CAN" = "#F8756D", "MEX" = "#33C051", "USA" = "#73A6FF")

##########################################################################################################################################
# GRAPH

gg_10d__facet <- ggplot(data_10d_core_melt, 
                        aes(
                          x = data_10d_core_melt$Border_Code,
                          y = value
                        )) +
  geom_col(
    #aes(fill = mode_color_code,color = Border_Code)
    # aes(fill = factor(data_10d_core_melt$col_fill_color)),
    aes(fill = factor(data_10d_core_melt$Border_Code)),
    color = "gray10",
    alpha = 1
  ) + 
  # geom_label(
  #   aes(label = scales::comma(data_10d_core_melt$value)
  #       # label = value # paste(round(crossings / 1e6, 0), "M")
  #   ),
  #   hjust = .5,
  #   vjust = -.5,
  #   fill = NA,
  #   label.size = 0
  # ) +
  geom_text(
    aes(label = scales::comma(data_10d_core_melt$value))
    ,vjust = -.8,
    ,size = 3.8
  ) + 
  # scale_fill_identity() +
  facet_wrap(vars(measure), nrow = 1, scales = "free_y")  +
  # facetted_pos_scales(
  #   y = list(
  #     vars(measure) == "Miles of Border" ~ scale_y_continuous(breaks = seq(0, 8000, by = 2000), limits= c(0,8000)),
  #     vars(measure) == "Points of Entry" ~ scale_y_continuous(breaks = seq(0, 120, by = 30), limits= c(0,120)),
  #     vars(measure) == "States on Border" ~ scale_y_continuous(breaks = seq(0, 16, by = 4), limits= c(0,16)),
  #     vars(measure) == "Border Crossings" ~ scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6), breaks = seq(0, 200000000, by = 50000000),limits= c(0,200000000))
  #   )
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0),
    axis.title.y = element_blank(),
    panel.background = element_blank(),
    strip.background = element_rect(colour = "gray50", fill = "gray90"),
    panel.border = element_rect(colour = "gray50", fill = NA, size = .2), 
    panel.grid.minor.y = element_blank(),        
    panel.grid.major.x = element_blank(),        
    panel.grid.minor.x = element_blank()
  ) +
  scale_x_discrete(
    name = "USA Border" #"data_10d_core_melt$Border_Code,
    # labels = unit_format(unit = "M", scale = 1e-6)
  ) +
  # scale_fill_manual(values =  data_10d_core_melt$col_fill_color) + #cols_border_crossing) +
  scale_fill_manual(values = cols_border_crossing) +  
  facetted_pos_scales(
    y = list(
      measure == "Miles of Border" ~ scale_y_continuous(breaks = seq(0, 8000, by = 2000),limits= c(0,8000)),
      measure == "Ports of Entry" ~ scale_y_continuous(breaks = seq(0, 120, by = 30),limits= c(0,120)), 
      measure == "States on Border" ~ scale_y_continuous(breaks = seq(0, 16, by = 4), limits= c(0,16)) ,
      measure == "Border Crossings" ~ scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6),breaks = seq(0, 200000000, by = 50000000),limits= c(0,200000000)) 
    )
  ) +
  labs(
    title = "USA International Borders | Overview",
    y = data_10d_core_melt$measure,
    subtitle = "Border Crossings is for 2023",
    caption = "Bureau of Transportation Statistics"
  )    + 
  guides(fill="none")

      
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align = 'left', fig.width = 10, fig.height=5}
  knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

#   suppressWarnings({ 
#     # 
# })   
  gg_00a_map_intl_borders
    # 
```

### US-Canada Border
The United States and **Canada** share the world's longest internation border, which is 5,525 miles long.  It spans 13 U.S states and has 84 ports of entry, with 1/3 of the northern border comprised of just the eastern flank of Alaska.

### US-Mexico Border
The United States border with **Mexico** is considerably shorter at 1,954 miles.  It covers 4 states with 28 ports of entry.  The **US-Mexico** border is about 1/3 as long, has 1/3 the Ports of Entry, covers 1/4 the number of states, but has 5X more border crossings than the **US-Canada** border.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align = 'left', fig.width = 10, fig.height=5}
  knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

#   suppressWarnings({ 
#     # 
# })  
  gg_10d__facet
    # 
```

### In-bound Crossing Volume and Density by Border
The US-Canada Border is 3X longer than the US-Mexico Border and has 3X more ports of entry to cover the distance.  In spite of the shorter border, the US-Mexico border has 4X more border crossings than the US-Canada Border. 

